---
title: "Mapping-Type Visualisation"
subtitle: "Many-to-one & many-to-many bipartite plots"
output: html_document
params:
  csv_path: "cBioPortal_disease_cancer_type_map.csv"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = TRUE,
  warning = FALSE,
  fig.width  = 12,
  fig.height = 8,
  dpi = 150
)
```

## 1 — Load libraries

```{r libraries}
library(ggplot2)
library(dplyr)
library(tidyr)
```

## 2 — Read the curation map

Set `csv_path` to any map CSV. You can change it here or pass it via the
`params` YAML header.

```{r read-data}
csv_path <- params$csv_path

# If the path doesn't exist as-is, look relative to the Rmd location
if (!file.exists(csv_path)) {
    csv_path <- file.path(
        dirname(knitr::current_input(dir = TRUE)),
        params$csv_path
    )
}
stopifnot(file.exists(csv_path))

message("Reading → ", csv_path)
dat <- read.csv(csv_path, stringsAsFactors = FALSE)
```

## 3 — Detect columns & delimiters

```{r detect}
col_names <- names(dat)

# Auto-detect the curated ontology column
if ("curated_ontology" %in% col_names) {
    curated_col <- "curated_ontology"
} else {
    candidates <- grep("[.:]curated_ontology$", col_names, value = TRUE)
    preferred  <- grep("^sample_type[.:]curated_ontology$", candidates,
                       value = TRUE)
    if (length(preferred) > 0) {
        curated_col <- preferred[1]
    } else if (length(candidates) > 0) {
        curated_col <- candidates[1]
    } else {
        stop("No curated_ontology column found in: ", csv_path)
    }
}
message("Curated column: ", curated_col)

# Normalise to two columns
dat_raw <- dat %>%
    select(original_value, curated_ontology = all_of(curated_col)) %>%
    filter(!is.na(curated_ontology), curated_ontology != "")

# Auto-detect delimiter
vals <- dat_raw$curated_ontology
if (any(grepl("<;>", vals, fixed = TRUE))) {
    delim <- "<;>"
} else if (any(grepl(";", vals, fixed = TRUE))) {
    delim <- ";"
} else {
    delim <- NULL
}
if (!is.null(delim)) message("Detected delimiter: '", delim, "'") else message("No delimiter detected")
```

## 4 — Expand & deduplicate

```{r expand}
if (!is.null(delim)) {
    delim_regex <- gsub("([.|()\\^{}+$*?\\[\\]])", "\\\\\\1", delim)
    dat_long <- dat_raw %>%
        mutate(row_id = row_number()) %>%
        separate_rows(curated_ontology, sep = delim_regex) %>%
        mutate(curated_ontology = trimws(curated_ontology)) %>%
        filter(curated_ontology != "") %>%
        select(original_value, curated_ontology) %>%
        distinct()
} else {
    dat_long <- dat_raw %>%
        mutate(curated_ontology = trimws(curated_ontology)) %>%
        select(original_value, curated_ontology) %>%
        distinct()
}

fwd <- dat_long %>%
    group_by(original_value) %>%
    summarise(n_curated = n_distinct(curated_ontology), .groups = "drop")

message("Unique original values: ", nrow(fwd))
message("  → map to 1 curated term: ", sum(fwd$n_curated == 1))
message("  → map to >1 curated term: ", sum(fwd$n_curated > 1))
```

## 5 — Identify many-to-one mappings

Multiple `original_value` entries that each map to the **same single**
`curated_ontology` term.

```{r many-to-one-data}
pure_one <- fwd %>% filter(n_curated == 1) %>% pull(original_value)

m2o_edges <- dat_long %>%
    filter(original_value %in% pure_one) %>%
    select(original_value, curated_ontology) %>%
    distinct()

m2o_groups <- m2o_edges %>%
    group_by(curated_ontology) %>%
    filter(n() > 1) %>%
    mutate(group_size = n()) %>%
    ungroup()

top3 <- m2o_groups %>%
    distinct(curated_ontology, group_size) %>%
    slice_max(group_size, n = 3) %>%
    pull(curated_ontology)

m2o_sel <- m2o_groups %>% filter(curated_ontology %in% top3)

message("Many-to-one groups found: ",
        n_distinct(m2o_groups$curated_ontology),
        " (showing top 3)")
```

## 6 — Identify many-to-many mappings

`original_value` entries that map to **more than one** curated term.

```{r many-to-many-data}
m2m_originals <- fwd %>% filter(n_curated > 1) %>% pull(original_value)

m2m_edges_all <- dat_long %>%
    filter(original_value %in% m2m_originals) %>%
    select(original_value, curated_ontology) %>%
    distinct()

m2m_originals_sel <- m2m_edges_all %>%
    group_by(original_value) %>%
    summarise(n = n(), .groups = "drop") %>%
    slice_max(n, n = 6) %>%
    pull(original_value)

m2m_sel <- m2m_edges_all %>% filter(original_value %in% m2m_originals_sel)

message("Many-to-many originals found: ", length(m2m_originals),
        " (showing top 6)")
```

## 7 — Define the plotting function

```{r plot-function}
pal_left  <- "#2b7bba"
pal_right <- "#e85d04"

edge_pal <- c("#6a4c93", "#1982c4", "#8ac926", "#ff595e", "#ff924c",
              "#56cfe1", "#457b9d", "#e63946", "#2a9d8f", "#f4a261")

bipartite_plot <- function(edges, title,
                           colour_by = c("right", "left"),
                           left_title  = "original_value",
                           right_title = "curated_ontology") {

    colour_by <- match.arg(colour_by)

    left_labs  <- unique(edges$original_value)
    right_labs <- unique(edges$curated_ontology)
    n_left     <- length(left_labs)
    n_right    <- length(right_labs)

    left_df <- data.frame(
        label = left_labs, x = 0,
        y = seq(1, 0, length.out = n_left),
        stringsAsFactors = FALSE
    )
    right_df <- data.frame(
        label = right_labs, x = 1,
        y = seq(1, 0, length.out = n_right),
        stringsAsFactors = FALSE
    )

    seg_df <- edges %>%
        left_join(left_df  %>% rename(x1 = x, y1 = y),
                  by = c("original_value"  = "label")) %>%
        left_join(right_df %>% rename(x2 = x, y2 = y),
                  by = c("curated_ontology" = "label"))

    if (colour_by == "right") {
        seg_df$colour_group <- seg_df$curated_ontology
    } else {
        seg_df$colour_group <- seg_df$original_value
    }
    colour_levels <- unique(seg_df$colour_group)
    colour_map <- setNames(
        rep_len(edge_pal, length(colour_levels)),
        colour_levels
    )

    ggplot() +
        geom_curve(
            data = seg_df,
            aes(x = x1, y = y1, xend = x2, yend = y2,
                colour = colour_group),
            curvature = 0.25, linewidth = 0.8, alpha = 0.6,
            show.legend = FALSE
        ) +
        scale_colour_manual(values = colour_map) +
        geom_label(
            data = left_df,
            aes(x = x - 0.02, y = y, label = label),
            hjust = 1, size = 2.4, fontface = "bold",
            colour = "white", fill = pal_left,
            linewidth = 0,
            label.r = unit(0.25, "lines"),
            label.padding = unit(0.2, "lines")
        ) +
        geom_label(
            data = right_df,
            aes(x = x + 0.02, y = y, label = label),
            hjust = 0, size = 2.4, fontface = "bold",
            colour = "white", fill = pal_right,
            linewidth = 0,
            label.r = unit(0.25, "lines"),
            label.padding = unit(0.2, "lines")
        ) +
        annotate("text", x = 0, y = 1.10, label = left_title,
                 fontface = "bold", size = 3.5, colour = "#444444") +
        annotate("text", x = 1, y = 1.10, label = right_title,
                 fontface = "bold", size = 3.5, colour = "#444444") +
        labs(title = title) +
        coord_cartesian(xlim = c(-0.75, 1.75), ylim = c(-0.06, 1.20),
                        clip = "off") +
        theme_void(base_family = "sans") +
        theme(
            plot.title      = element_text(face = "bold", hjust = 0.5,
                                           size = 15, margin = margin(b = 6)),
            plot.background = element_rect(fill = "#fafafa", colour = NA),
            plot.margin     = margin(12, 12, 12, 12)
        )
}
```

## 8 — Many-to-One plot

```{r plot-m2o, fig.width=12, fig.height=15}
if (nrow(m2o_sel) > 0) {
    bipartite_plot(m2o_sel, "Many-to-One Mapping", colour_by = "right")
} else {
    message("No many-to-one mappings to display.")
}
```

## 9 — Many-to-Many plot

```{r plot-m2m, fig.width=12, fig.height=8}
if (nrow(m2m_sel) > 0) {
    bipartite_plot(m2m_sel, "Many-to-Many Mapping", colour_by = "left")
} else {
    message("No many-to-many mappings to display.")
}
```

---

*Generated from `plot_mapping_types.Rmd`*
