---
title: "Attribute Name Heterogeneity: cBioPortal & curatedMetagenomicData"
author: Sehyun Oh
date: "`r format(Sys.time(), '%B %d, %Y')`"
format:
    html:
        fontsize: 14px
        toc: true
        toc-depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, collapse = TRUE)
```

```{r libs, echo=FALSE}
suppressPackageStartupMessages({
    library(dplyr)
    library(readr)
    library(ggplot2)
    library(knitr)
    library(kableExtra)
})
```

## Overview

A single biological concept — *age at sample collection*, *sex*, *disease diagnosis* — is
routinely recorded under dozens of different column names across studies in the same database.
This document quantifies that naming heterogeneity for two large biomedical data resources:

- **cBioPortal** – cancer genomics (clinical metadata aggregated from hundreds of studies)
- **curatedMetagenomicData (cMD)** – human microbiome (manually curated from public shotgun studies)

For each harmonized concept we count the number of **distinct original column names** that
contributors used to encode the same information. These counts come directly from the
merging schemas built during the OmicsMLRepo curation process.

---

## Load merging schemas

```{r load}
extDir <- "~/OmicsMLRepo/OmicsMLRepoData/inst/extdata"

cbio_ms <- read_csv(file.path(extDir, "cBioPortal_merging_schema.csv"),
                    show_col_types = FALSE)
cmd_ms  <- read_csv(file.path(extDir, "cMD_merging_schema.csv"),
                    show_col_types = FALSE)
```

---

## Parse attribute name counts

```{r parse}
# Helper: assign a broad semantic category to a curated field name
assign_category <- function(field) {
    case_when(
        grepl("^age", field)                                                    ~ "Age",
        grepl("^treatment|antibiotics|dietary", field)                          ~ "Treatment",
        grepl("cancer|disease|tumor|vital|staging|subtype|specimen|sample_type|body_site|feces|disease_response|hla|biomarker|bmi", field) ~ "Disease / Biology",
        grepl("ancestry|sex|country|location|westernized|smoker|obgyn|neonatal|fmt|pmid|ncbi|dna_extract|probing", field) ~ "Demographics / Sample",
        TRUE                                                                    ~ "Other"
    )
}

# ── cBioPortal ──────────────────────────────────────────────────────────────
cbio_dat <- cbio_ms %>%
    # exclude rows where no original field exists (newly introduced concepts)
    filter(!is.na(original_field), !is.na(original_field_num)) %>%
    mutate(
        # take the first curated field name when multiple are joined with ";"
        concept      = gsub(";.*", "", curated_field),
        n_original   = as.integer(original_field_num),
        dataset      = "cBioPortal",
        category     = assign_category(concept)
    ) %>%
    select(concept, n_original, dataset, category)

# ── curatedMetagenomicData ───────────────────────────────────────────────────
cmd_dat <- cmd_ms %>%
    filter(!is.na(original_field), original_field != "NA",
           !is.na(original_field_num)) %>%
    mutate(
        concept      = gsub(";.*", "", curated_field),
        n_original   = as.integer(original_field_num),
        dataset      = "cMD",
        category     = assign_category(concept)
    ) %>%
    select(concept, n_original, dataset, category)

combined <- bind_rows(cbio_dat, cmd_dat)
```

---

## Summary table {#sec-table}

```{r summary_table}
summary_tbl <- combined %>%
    group_by(dataset) %>%
    summarise(
        `Harmonized concepts`            = n(),
        `Total original attribute names` = sum(n_original),
        `Median names / concept`         = median(n_original),
        `Max names / concept`            = max(n_original),
        `Concepts with ≥ 10 names`       = sum(n_original >= 10),
        .groups = "drop"
    )

kable(summary_tbl,
      caption = "**Table 1.** Attribute name heterogeneity across two biomedical databases.
      Each 'original attribute name' is a distinct column header found across
      contributing studies before harmonization.",
      align = c("l", rep("r", ncol(summary_tbl) - 1))) %>%
    kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

---

## Figure: Distribution of attribute name synonyms {#sec-figure}

```{r heterogeneity_fig, fig.width=9, fig.height=6.5, dpi=300}
## ── colour palette ───────────────────────────────────────────────────────
pal <- c(
    "Age"                   = "#E64B35",
    "Treatment"             = "#F39B7F",
    "Disease / Biology"     = "#4DBBD5",
    "Demographics / Sample" = "#00A087",
    "Other"                 = "#B09C85"
)

## ── build plot data ───────────────────────────────────────────────────────
# Create a unique label per concept-dataset combination, then set its factor
# level order by n_original so each facet panel is sorted independently.
plot_dat <- combined %>%
    # unique key: concept + dataset avoids level collisions across facets
    mutate(key = paste0(concept, " [", dataset, "]")) %>%
    arrange(dataset, n_original) %>%
    mutate(key = factor(key, levels = unique(key)))

## ── lollipop / Cleveland dot plot ────────────────────────────────────────
p <- ggplot(plot_dat, aes(x = n_original, y = key, colour = category)) +
    geom_segment(aes(xend = 0, yend = key),
                 linewidth = 0.5, alpha = 0.45) +
    geom_point(size = 2.8) +
    scale_x_log10(
        breaks = c(1, 3, 10, 30, 100, 300),
        labels = c("1", "3", "10", "30", "100", "300")
    ) +
    # strip " [cBioPortal]" / " [cMD]" from y-axis tick labels
    scale_y_discrete(labels = function(x) sub(" \\[.*\\]$", "", x)) +
    scale_colour_manual(values = pal, name = "Semantic category") +
    facet_wrap(~ dataset, scales = "free_y", ncol = 2) +
    theme_bw(base_size = 11) +
    labs(
        title    = "Naming heterogeneity of metadata attributes in cBioPortal and curatedMetagenomicData",
        subtitle = paste0(
            "Each dot = one harmonized concept; x-axis = # of distinct original column names ",
            "(log\u2081\u2080 scale).\n",
            "Multiple contributing studies use different column names for the same biological concept."
        ),
        x = "# of distinct original attribute names (log\u2081\u2080 scale)",
        y = NULL
    ) +
    theme(
        axis.text.y        = element_text(size = 7.5),
        strip.text         = element_text(face = "bold", size = 11),
        legend.position    = "bottom",
        legend.title       = element_text(size = 9),
        legend.text        = element_text(size = 8),
        plot.title         = element_text(face = "bold", size = 11),
        plot.subtitle      = element_text(size = 9, colour = "grey40"),
        panel.grid.major.y = element_line(linetype = "dotted", colour = "grey85")
    )

print(p)
```

```{r save_fig}
ggsave("figures/attribute_heterogeneity.png",
       plot = p, width = 9, height = 6.5, dpi = 600)
```

---

## Key callout examples {#sec-callout}

```{r age_example}
# cBioPortal age-related fields
age_cbio <- cbio_ms %>%
    filter(grepl("^age_", curated_field),
           !is.na(original_field_num)) %>%
    select(curated_field, original_field_num, original_field)

# How many distinct original names encode "age" across all age-related curated fields?
total_age_names_cbio <- sum(as.integer(age_cbio$original_field_num), na.rm = TRUE)

cat("cBioPortal – age-related curated fields:", nrow(age_cbio), "\n")
cat("cBioPortal – total original 'age' column names:", total_age_names_cbio, "\n\n")

# A sample of original age names from age_at_procurement (most heterogeneous)
age_proc_names <- age_cbio %>%
    filter(curated_field == "age_at_procurement") %>%
    pull(original_field) %>%
    strsplit(";") %>%
    unlist()

cat("Example: original column names for 'age at procurement' in cBioPortal:\n")
cat(paste(" •", head(age_proc_names, 12), collapse = "\n"), "\n")
```

```{r treatment_example}
trt_cbio <- cbio_ms %>%
    filter(grepl("^treatment", curated_field),
           !is.na(original_field_num)) %>%
    select(curated_field, original_field_num) %>%
    mutate(original_field_num = as.integer(original_field_num))

cat("cBioPortal – treatment-related harmonized fields:", nrow(trt_cbio), "\n")
cat("cBioPortal – total original 'treatment' column names:",
    sum(trt_cbio$original_field_num), "\n")
cat("\nBreakdown by treatment sub-concept:\n")
print(trt_cbio)
```

---

## Interpretation

The table and figure above demonstrate that **naming inconsistency is pervasive and
substantial** even within a single curated database:

- In **cBioPortal**, the 30 harmonized clinical concepts are represented by a total of
  **`r sum(cbio_dat$n_original)` distinct original column names** across contributing studies.
  A single concept such as *age* is encoded under
  **`r total_age_names_cbio` different column names** (e.g., `AGE`, `AGE_IN_DAYS`,
  `AGE_AT_PROCUREMENT`, `AGE_AT_SEQ_REPORT`, `PATIENT_AGE_AT_DIAGNOSIS`, …), and
  *treatment type* alone attracts **`r cbio_dat %>% filter(concept=='treatment_type') %>% pull(n_original)` names**.

- In **curatedMetagenomicData**, the naming is more standardised — the median concept has
  just **`r median(cmd_dat$n_original)` original name(s)** — yet *biomarker* still
  aggregates **`r cmd_dat %>% filter(concept=='biomarker') %>% pull(n_original)` original
  column names** from the individual study data tables.

This heterogeneity motivates an AI/ML-assisted approach to metadata harmonization that can
learn the mapping from heterogeneous raw attribute names to a consistent, ontology-grounded
schema at scale.

---

## For the review article

cBioPortal only, with treatment_* attributes excluded.

```{r review_data}
cbio_no_trt <- cbio_dat %>%
    filter(!grepl("^treatment", concept))
```

### Summary table (cBioPortal, no treatment)

```{r review_table}
review_tbl <- cbio_no_trt %>%
    summarise(
        Dataset                          = "cBioPortal (non-treatment)",
        `Harmonized concepts`            = n(),
        `Total original attribute names` = sum(n_original),
        `Median names / concept`         = median(n_original),
        `Max names / concept`            = max(n_original),
        `Concepts with ≥ 10 names`       = sum(n_original >= 10)
    )

kable(review_tbl,
      caption = "**Table 2.** Attribute name heterogeneity in cBioPortal (treatment attributes excluded).",
      align = c("l", rep("r", ncol(review_tbl) - 1))) %>%
    kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

### Figure (cBioPortal, no treatment) {#sec-review-fig}

```{r review_fig, fig.width=8, fig.height=5.5, dpi=300}
review_plot_dat <- cbio_no_trt %>%
    filter(n_original > 1) %>%
    arrange(n_original) %>%
    mutate(concept = factor(concept, levels = unique(concept)))

p_review <- ggplot(review_plot_dat,
                   aes(x = n_original, y = concept, colour = category)) +
    geom_segment(aes(xend = 0, yend = concept),
                 linewidth = 0.5, alpha = 0.45) +
    geom_point(size = 3) +
    scale_x_log10(
        breaks = c(1, 3, 10, 30, 100),
        labels = c("1", "3", "10", "30", "100"),
        limits = c(1, NA)
    ) +
    scale_colour_manual(values = pal, name = "Semantic category") +
    theme_bw(base_size = 11) +
    labs(
        title    = "Attribute name heterogeneity – cBioPortal",
        subtitle = paste0(
            "Each dot = one harmonized concept (treatment attributes excluded).\n",
            "x-axis = # of distinct original column names (log\u2081\u2080 scale)."
        ),
        x = "# of distinct original attribute names (log\u2081\u2080 scale)",
        y = NULL
    ) +
    theme(
        axis.text.y        = element_text(size = 9),
        legend.position    = "bottom",
        legend.title       = element_text(size = 9),
        legend.text        = element_text(size = 8),
        plot.title         = element_text(face = "bold", size = 11),
        plot.subtitle      = element_text(size = 9, colour = "grey40"),
        panel.grid.major.y = element_line(linetype = "dotted", colour = "grey85")
    )

print(p_review)
```

```{r review_fig_save}
ggsave("figures/attribute_heterogeneity_cbio_no_trt.png",
       plot = p_review, width = 6.5, height = 5.5, dpi = 600)
```

---

## Session Info
<details>
```{r session_info}
sessionInfo()
```
</details>
